WASP-4 is Accelerating Towards the Sun

### 2020-02-09 paper/ms.tex
[WIP] tweaks to JNW merge

@@

%TODO FIXME
\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle image.
  We expect that 1 out of 6 hot Jupiters will show apparent period
  changes of the same order of magnitude as WASP-4, based on the
  hot-Jupiter companion statistics of  \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

The orbits of most hot Jupiters are formally unstable to tidal decay
\citep{counselman_outcomes_1973,
hut_stability_1980,rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
It is not clear, though, whether the timescale for tidal orbital decay
is shorter or longer than the timescale for main-sequence stellar
evolution.  This answer to this question depends on the uncertain rate
at which friction inside the star damps the tidal oscillations, which
is ultimately what causes the orbit to shrink (as reviewed by
\citealt{Mazeh2008} and \citealt{ogilvie_tidal_2014}).  Population
studies of hot Jupiters --- based on ages, rotation rates, orbital
distances, and Galactic kinematics --- have led to differing
conclusions, with estimated lifetimes ranging from less than a
gigayear to much longer than main-sequence lifetimes (see, {\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).

@@

The focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  The period decrease was identified by
\citet{bouma_wasp4b_2019}]{bouma_wasp4b_2019}, who combined data from
the NASA TESS mission \citep{ricker_transiting_2015} and a decade of
ground-based transit observations. Soon thereafter,
\citet{southworth_transit_2019} reported an additional 22 transit
times and recalculated the period derivative to be $\dot{P} = -9.2 \pm
1.1$ milliseconds per year.  A separate study by \citet{baluev_2019}
reported on additional transit times, and pointed out that the period
decrease was statistically significant only when analyzing the data
with the highest precision.

To determine the origin of the period change, we acquired four
additional radial velocity measurements with the Keck I 10m telescope
and the High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}).  In doing so, we extended the time
baseline of HIRES measurements from 3 to 9 years.  The previously
available HIRES data led to the marginal ($\approx$$2\sigma$)
detection of a radial-velocity trend \citep{knutson_friends_2014}.
Our new measurements reveal a line-of-sight acceleration of
$\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
m\,s^{-1}\,day^{-1}}$.  Through the Doppler effect\footnote{While the
apparent period change caused by a line-of-sight acceleration has been
referred to as the ``R{\o}mer effect'' \citep{yee_orbit_2020}, a
simpler and probably better term is the Doppler effect. The R{\o}mer
effect refers to the {\it delay} in the reception of a signal due to a
change in the time required for light to traverse the distance between
the source and observer. The Doppler effect refers to the change in
the apparent {\it rate} of a process due to changes in the relative
motion of the source and observer, such as the rate of transits.},
this translates into an expected period decrease $-5.9$ milliseconds
per year, which is comparable to the period decrease that was measured
from transit timing.  We undertook high-resolution (speckle) imaging
to search for evidence of a companion that could be responsible for
the acceleration of WASP-4.

@@

% FIXME : update wording
Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.
% Five of the twenty-two occurred at simultaneous epochs as transits
% previously reported by \citet{hoyer_tramos_2013}.
Additional timing measurements were also reported recently by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.  We included twelve of their ``high
quality'' transit times from the TRAPPIST telescope (six transits),
El~Sauce (four transits), and \citet{petrucci_no_2013}.  For TRAPPIST
and El~Sauce, we verified with the original observers that correct
barycentric and leap-second corrections had been performed (M.~Gillon,
P.~Evans, priv{.} comm{.}).  We omitted the fourteen remaining
\citeauthor{baluev_2019} ETD\footnote{\url{http://var2.astro.cz/ETD}}
times due to ambiguity in whether leap-second corrections had been
performed.  We did not include in our analysis the four occultation
times tabulated by \citetalias{bouma_wasp4b_2019}, because of the
large timing uncertainties and negligible statistical power.

@@

The complete set of radial velocity observations is given in Table~2.
Along with the 2010--2019 HIRES observations are early measurements
with two different spectrographs.  We included data from CORALIE ({\bf
Spell it out on first use)} from \citet{wilson_wasp-4b_2008} and
\citet{triaud_spin-orbit_2010}, adopting the values from the
homogeneous analysis of the latter authors.  We included data from
HARPS ({\bf spell it out}) from \citet{pont_determining_2011} and
\citet{husnoo_observational_2012}.  While
\citet{triaud_spin-orbit_2010} also acquired HARPS data over three
nights for Rossiter-McLaughlin observations, these data were reduced
with a non-standard pipeline making them ill-suited for our study, and
we did not include them.\footnote{This problem was fixed in principle
by \citet{trifonov_public_2020} who performed a homogeneous
re-reduction of the entire HARPS data archive.  We found that the
decision regarding whether to include or omit these points did not
noticeably affect our results.}

@@

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times
60$$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic
interference pattern. This pattern in then used to determine the
properties of the detected companion and produce a reconstructed
image.  Using the reconstructed speckle images, contrast curves are
produced to determine the 5-$\sigma$ detection limits (see
\citealt{howell_speckle_2011}).  No companions were detected. The data
from the second night, which had better seeing (0.6$''$), led to the
most constraining limit.  The 832$\,$nm limits were the most useful,
given that any faint companion would likely be redder than the host
star.  Therefore, we opted to use the 832$\,$nm September 28-29
contrast limits in the analysis described below.

@@

We fitted two simple timing models to the transit-timing data. The
first model assumes the period $P$ to be constant:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the integer specifying the number within the sequence of
orbits spanned by the data, and $t_0$ is the transit time for the
event designated $E=0$.  The second model assumes that the period
changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference time $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the times of observed transits.
This choice leads to a small covariance between the uncertainties in
$t_0$ and $P$.
%A third possible
%model that we did not consider for reasons that will become apparent
%is a precessing, eccentric orbit \citep[{\it
%e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

Our initial model for the radial velocity data was a circular orbit
plus instrument offsets, ``jitter'' values (explained below), and a
long-term linear trend \citep[][\texttt{radvel}]{fulton_radvel_2018}.
We set Gaussian priors on the orbital period and time of inferior
conjunction using the values from Table~4 of
\citetalias{bouma_wasp4b_2019}.  We assumed the orbit to be circular
based on previous studies that placed stringent upper limits on the
eccentricity
\citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the orbital velocity semi-amplitude, the
instrument zero-points, an additive ``white noise'' instrument jitter
for each instrument, and a linear ($\dot{v_{\rm r}}$) acceleration
term.  (Without the linear trend, the model is a much poorer fit with
$\Delta {\rm BIC}=73$.)

@@

Because of the Doppler effect, any line-of-sight acceleration should
lead to an change in the observed orbital period:
\begin{align}
  \dot{P}_{\rm\,RV} &= \frac{\dot{v}_{\rm r} P}{c},
\end{align}
or in more convenient units,
\begin{align}
  \dot{P}_{\rm\,RV} &= 105{.}3\,{\rm msec}\,{\rm yr}^{-1} \
  \left( \frac{P}{{\rm day}} \right)
  \left( \frac{\dot{\gamma}}{ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}} \right).
\end{align}
For WASP-4, this yields
\begin{align}
  \dot{P}_{\rm\,RV} &= -5.94 \pm 0.39\,{\rm msec}\,{\rm yr}^{-1}.
\end{align}
Therefore, most of the period derivative that was detected through
transit timing ($\dot{P}= - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$)
can be accounted for by the Doppler effect.  Given the evidence for
unmodeled noise in both the transit timing data and the
radial-velocity data, it seems plausible that the Doppler effect can
account for the entire observed period derivative.

@@

For each grid cell, we simulated 512 hypothetical companions. We
assigned each companion a mass and semimajor axis from log-uniform
distributions within the boundaries of the grid cell. We drew the
inclination from a uniform distribution in $\cos i$.  For companion
masses less than $10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity from the power-law $p_e \propto
e^\eta$ reported by \citet{moe_mind_2017} in their Equation~17 ($\eta
\approx 0.5$ for most orbital periods).  The long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero.  The binary star distribution is
``top-heavy'', with a broad range of eccentricities extending close to
unity.  The choice of $10\,M_{\rm Jup}$ as the dividing line between
these two regimes was based on the empirical study of
\citet{schlaufman_evidence_2018} on the distinction between giant
planets and brown dwarfs. This value is also close to the $13\,M_{\rm
Jup}$ deuterium-burning limit \citep[{\it
e.g.},][]{burrows_nongray_1997}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: make table 4 look beter

@@

%TODO FIXME: some sentences
\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle image.
  We expect that 1 out of 6 hot Jupiters will show apparent period
  changes of the same order of magnitude as WASP-4, based on the
  hot-Jupiter companion statistics of  \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

The orbits of most hot Jupiters are formally unstable to tidal decay
\citep{counselman_outcomes_1973,hut_stability_1980,rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
It is not clear, though, whether the timescale for tidal orbital decay
is shorter or longer than the timescale for main-sequence stellar
evolution.  This answer to this question depends on the uncertain rate
at which friction inside the star damps the tidal oscillations, which
is ultimately what causes the orbit to shrink (as reviewed by
\citealt{Mazeh2008} and \citealt{ogilvie_tidal_2014}).  Population
studies of hot Jupiters --- based on ages, rotation rates, orbital
distances, and Galactic kinematics --- have led to differing
conclusions, with estimated lifetimes ranging from less than a
gigayear to much longer than main-sequence lifetimes (see, {\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).

@@

The focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  The period decrease was identified by \citep[][hereafter
\citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}, who combined data
from the NASA TESS mission \citep{ricker_transiting_2015} and a decade
of ground-based transit observations. Soon thereafter,
\citet{southworth_transit_2019} reported an additional 22 transit
times and recalculated the period derivative to be $\dot{P} = -9.2 \pm
1.1$ milliseconds per year.  A separate study by \citet{baluev_2019}
reported on additional transit times, and pointed out that the period
decrease was statistically significant only when analyzing the data
with the highest precision.

To determine the origin of the period change, we acquired four
additional radial velocity measurements with the Keck I 10m telescope
and the High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}).  In doing so, we extended the time
baseline of HIRES measurements from 3 to 9 years.  The previously
available HIRES data led to the marginal ($\approx$$2\sigma$)
detection of a radial-velocity trend \citep{knutson_friends_2014}.
Our new measurements reveal a line-of-sight acceleration of
$\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
m\,s^{-1}\,day^{-1}}$.  Through the Doppler effect\footnote{While the
apparent period change caused by a line-of-sight acceleration has been
referred to as the ``R{\o}mer effect'' \citep{yee_orbit_2020}, a
simpler and probably better term is the Doppler effect. The R{\o}mer
effect refers to the {\it delay} in the reception of a signal due to a
change in the time required for light to traverse the distance between
the source and observer. The Doppler effect refers to the change in
the apparent {\it rate} of a process due to changes in the relative
motion of the source and observer, such as the rate of transits.},
this translates into an expected period decrease of $-5.9$
milliseconds per year, which is comparable to the period decrease that
was measured from transit timing.  We undertook high-resolution
(speckle) imaging to search for evidence of a companion that could be
responsible for the acceleration of WASP-4.

@@

Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.  Additional timing measurements were also
reported recently by \citet{baluev_2019}, based on a homogeneous
analysis of archival ground-based observations.  We included twelve of
their ``high quality'' transit times from the TRAPPIST telescope (six
transits), El~Sauce (four transits), and \citet{petrucci_no_2013}.
For TRAPPIST and El~Sauce, we verified with the original observers
that the timestamps were amenable to the appropriate barycentric and
leap second corrections (M.~Gillon, P.~Evans, priv{.} comm{.}).  We
omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had been performed.  We did not
include in our analysis the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

The complete set of radial velocity observations is given in Table~2.
Along with the 2010--2019 HIRES observations are early measurements
with two different spectrographs.  \citet{wilson_wasp-4b_2008} and
\citet{triaud_spin-orbit_2010} observed WASP-4 with the Swiss 1.2m
Euler Telescope and CORALIE Spectrograph; we adopted the radial
velocity values from the homogeneous analysis of the latter authors.
We also included data from the High Accuracy Radial Velocity Planet
Searcher (HARPS), reported by \citet{pont_determining_2011} and
\citet{husnoo_observational_2012}.  While
\citet{triaud_spin-orbit_2010} also acquired HARPS data over three
nights for Rossiter-McLaughlin observations, these data were reduced
with a non-standard pipeline making them ill-suited for our study, and
we did not include them.\footnote{This problem was fixed in principle
by \citet{trifonov_public_2020} who performed a homogeneous
re-reduction of the entire HARPS data archive.  We found that the
decision regarding whether to include or omit these points did not
noticeably affect our results.}

@@

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times
60$$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic
interference pattern. This pattern is then used to determine the
properties of the detected companion and produce a reconstructed
image.  Using the reconstructed speckle images, contrast curves are
produced to determine the 5-$\sigma$ detection limits (see
\citealt{howell_speckle_2011}).  No companions were detected. The data
from the second night, which had better seeing (0.6$''$), led to the
most constraining limit.  The 832$\,$nm limits were the most useful,
given that any faint companion would likely be redder than the host
star.  Therefore, we opted to use the 832$\,$nm September 28-29
contrast limits in the analysis described below.

@@

We fitted two simple timing models to the transit-timing data. The
first model assumes the period $P$ to be constant:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the integer specifying the number within the sequence of
orbits spanned by the data, and $t_0$ is the transit time for the
event designated $E=0$.  The second model assumes that the period
changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference time $t_0$, the period at the
reference time $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the times of observed transits.
This choice leads to a small covariance between the uncertainties in
$t_0$ and $P$.

@@

Our initial model for the radial velocity data was a circular orbit
plus instrument offsets, ``jitter'' values (explained below), and a
long-term linear trend \citep[][\texttt{radvel}]{fulton_radvel_2018}.
We set Gaussian priors on the orbital period and time of inferior
conjunction using the values from Table~4 of
\citetalias{bouma_wasp4b_2019}.  We assumed the orbit to be circular
based on previous studies that placed stringent upper limits on the
eccentricity
\citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the orbital velocity semi-amplitude, the
instrument zero-points, a ``white noise'' instrument jitter for each
instrument added in quadrature to its uncertainties, and a linear
($\dot{v_{\rm r}}$) acceleration term.  (Without the linear trend, the
model is a much poorer fit with $\Delta {\rm BIC}=73$.)

@@

Because of the Doppler effect, any line-of-sight acceleration should
lead to a change in the observed orbital period:
\begin{align}
  \dot{P}_{\rm\,RV} &= \frac{\dot{v}_{\rm r} P}{c},
\end{align}
or in more convenient units,
\begin{align}
  \dot{P}_{\rm\,RV} &= 105{.}3\,{\rm msec}\,{\rm yr}^{-1} \
  \left( \frac{P}{{\rm day}} \right)
  \left( \frac{\dot{\gamma}}{ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}} \right).
\end{align}
For WASP-4, this yields
\begin{align}
  \dot{P}_{\rm\,RV} &= -5.94 \pm 0.39\,{\rm msec}\,{\rm yr}^{-1}.
\end{align}
Therefore, most of the period derivative that was detected through
transit timing ($\dot{P}= - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$)
can be accounted for by the Doppler effect.  Given the evidence for
unmodeled noise in both the transit timing data and the
radial-velocity data, it seems plausible that the Doppler effect can
account for the entire observed period derivative.

@@

For each grid cell, we simulated 512 hypothetical companions. We
assigned each companion a mass and semimajor axis from log-uniform
distributions within the boundaries of the grid cell. We drew the
inclination from a uniform distribution in $\cos i$.  For companion
masses less than $10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity from the power-law $p_e \propto
e^\eta$ reported by \citet{moe_mind_2017} in their Equation~17 ($\eta
\approx 0.5$ for most orbital periods).  The long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero.  The binary star distribution is
``top-heavy'', with a broad range of eccentricities extending close to
unity \citep{moe_mind_2017,price-whelan_close_2020}.  The choice of
$10\,M_{\rm Jup}$ as the dividing line between these two regimes was
based on the empirical study of \citet{schlaufman_evidence_2018} on
the distinction between giant planets and brown dwarfs. This value is
also close to the $13\,M_{\rm Jup}$ deuterium-burning limit
\citep[{\it e.g.},][]{burrows_nongray_1997}.
